Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science

### 2025-01-30 paper.qmd
up to the conclusion

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@@

Poor communication about dependencies is widespread in code written by researchers in many disciplines, and a common source of irreproducible results [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

@@

Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to efforts that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several ways. First, authors should use a clear and logical folder structure for their code and data. There are many simple templates to guide authors on this, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, \@Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine, before submission to confirm everything works as expected. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets]. Third, authors should include in their compendium a README document that describes to a reader the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper. A simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Archaeologists are using these tools write complete journal articles, e.g. @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024.

Dependencies

@@

so efforts aimed at systemic improvements of dependency handling—for example, as per [134]—have great potential to increase reproducibility.... it is hard to assess whether reported findings are credible.

>>>>

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by a generous measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@@

@@

Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a clear and logical folder structure for their code and data. There are many simple templates to guide authors on this, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine (e.g. by another co-author of the paper), before submission to confirm everything works as expected. This would likely also prevent issues relating to path and function errors. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets]. Third, authors should include in their compendium a README document that describes to a reader the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper. A simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscript using Quarto. These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.

Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025]. The simplest way for archaeologists, to improve here is to write the names and version numbers of the software and packages they used in their README file, as we see in @herskindComputationalLinguisticMethodology2024 and @monnaStudyingSerialityMaterial2024. For more complex research projects, i.e. those using five or more packages or machine learning algorithms, authors should use dependency management tools to keep track of the packages and version numbers needed to reproduce their results. This is an active area of development, and while there are many tools currently available, the most robust and widely used include renv for R [@renv, see examples in @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024] and conda and poetry for Python [@conda2023; @poetry2023]. A more comprehensive solution, and the leading best practice for managing dependencies in many computationally intensive fields using Python in particular, is to include a Dockerfile in the compendium [@moreauContainersComputationalReproducibility2023]. This is a set of a machine- and human-readable instructions that enables a user to recreate the author's computational environment (including those requirements beyond the R or Python packages) on another computer [@nustTenSimpleRules2020]. Dockerfiles are gradually being adopted by archaeologists, see @cremaModellingDiffusionInnovation2024 and @liaoApplyingMaskRCNN2024 for examples, and most of my reproducibility reviews include a recommendation that the author include a Dockerfile to manage dependencies efficiently.

## Conclusion

@@

limitations: not looked at FAIR or CARE attributes of data, not looked at code quality (efficiency, readability, style)
